Possible Inconsistency in Maximum Likelihood Calculation of Ancestral States

### 2017-03-17 main.tex
update

%% Useful packages
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}

@@

Define the ``Farris zone'' topology $\tau_1$ as \cite{farris}
\begin{verbatim}
1\ /3
  | 
2/ \4
\end{verbatim}
and the ``Felsenstein zone'' topology $\tau_2$ as \cite{felsenstein}
\begin{verbatim}
1\      /3
   >--< 
2/      \4
\end{verbatim}
where, for generating data $t^*$, the probability of a change in character along the top two branches and the middle branch is $p_x$ and the probability of a change in character along the bottom two branches is $p_y$.
In our estimand $t$, we modify the classical topologies by allowing the middle branch to be unequal to either $p_x$ or $p_y$ and denote this probability $p_w$; the parameters $p_x, p_y$ and $p_w$ all lie in interval $[0,1/2]$.
To obtain branch length parameters, we transform these probabilities using the function $f(p)=1-2p$ so that $t=\{x,y,w\}:=\{f(p_x), f(p_y), f(p_w)\}$; the parameters $x, y$ and $w$ all lie in the interval $[0,1]$.

@@

\begin{table}
\centering
\begin{tabular}{|l|l|l|}
\multicolumn{3}{c}{Classical}\\
    \hline
$s_j$   &$P(s_j|\tau_1,t)$&$P(s_j|\tau_2,t)$\\
    \hline
0000&$1+x^2+y^2+4xy^2+x^2y^2$&$1+2xy+2xy^2+x^2y+y^3+x^2y^2$\\
0001&$1+x^2-y^2-x^2y^2$&$1+x^2y-y^3-x^2y^2$\\
0010&$1-x^2+y^2-x^2y^2$&$1-x^2y+y^3-x^2y^2$\\
0100&$1+x^2-y^2-x^2y^2$&$1+x^2y-y^3-x^2y^2$\\
1000&$1-x^2+y^2-x^2y^2$&$1-x^2y+y^3-x^2y^2$\\
0011&$1-x^2-y^2+x^2y^2$&$1+2xy-2xy^2-x^2y-y^3+x^2y^2$\\
0101&$1+x^2+y^2-4xy^2+x^2y^2$&$1-2xy-2xy^2+x^2y+y^3+x^2y^2$\\
1001&$1-x^2-y^2+x^2y^2$&$1-2xy+2xy^2-x^2y-y^3+x^2y^2$\\
    \hline
\multicolumn{3}{c}{Modified}\\
    \hline
$s_j$   &$P(s_j|\tau_1,t)$&$P(s_j|\tau_2,t)$\\
    \hline
0000&$1+x^2+w^2+4xyw+x^2w^2$&$1+2xw+2xyw+x^2y+w^2y+x^2w^2$\\
0001&$1+x^2-w^2-x^2w^2$&$1+x^2y-w^2y-x^2w^2$\\
0010&$1-x^2+w^2-x^2w^2$&$1-x^2y+w^2y-x^2w^2$\\
0100&$1+x^2-w^2-x^2w^2$&$1+x^2y-w^2y-x^2w^2$\\
1000&$1-x^2+w^2-x^2w^2$&$1-x^2y+w^2y-x^2w^2$\\
0011&$1-x^2-w^2+x^2w^2$&$1+2xw-2xyw-x^2y-w^2y+x^2w^2$\\
0101&$1+x^2+w^2-4xyw+x^2w^2$&$1-2xw-2xyw+x^2y+w^2y+x^2w^2$\\
1001&$1-x^2-w^2+x^2w^2$&$1-2xw+2xyw-x^2y-w^2y+x^2w^2$\\
    \hline
\end{tabular}    
\caption{Site pattern probabilities.
All values are multiplied by $1/8$.}
\label{tab:sitepatprob}
\end{table}

\begin{table}
\centering
\begin{tabular}{|l|ll|}
\multicolumn{3}{c}{$P(z_j|s_j,\tau_1,t)$}\\
\hline
& \multicolumn{2}{|c|}{$z_j$}\\
    \hline
$s_j$    &00                              &01\\
    \hline
0000&$(1+x)^2(1+y)(1+w)^{2*}$          &$(1+x)^2(1-y)(1-w)^2$\\
0001&$(1+x)^2(1+y)(1+w)(1-w)^*$       &$(1+x)^2(1-y)(1+w)(1-w)$\\
0010&$(1+x)^2(1+y)(1+w)(1-w)^*$       &$(1+x)^2(1-y)(1+w)(1-w)$\\
0100&$(1+x)(1-x)(1+y)(1+w)^{2*}$       &$(1+x)(1-x)(1-y)(1-w)^2$\\
1000&$(1+x)(1-x)(1+y)(1+w)^{2*}$       &$(1+x)(1-x)(1-y)(1-w)^2$\\
0011&$(1+x)(1-x)(1+y)(1+w)(1-w)^{\dagger}$    &$(1+x)(1-x)(1-y)(1+w)(1-w)$\\
0101&$(1+x)^2(1+y)(1-w)^{2\ddagger}$          &$(1+x)^2(1-y)(1+w)^{2\ddagger}$\\
1001&$(1+x)(1-x)(1+y)(1+w)(1-w)^{\dagger}$    &$(1+x)(1-x)(1-y)(1+w)(1-w)$\\
    \hline
    \hline
&10                           &11\\
    \hline
0000&$(1-x)^2(1-y)(1+w)^2$        &$(1-x)^2(1+y)(1-w)^2$\\
0001&$(1-x)^2(1-y)(1+w)(1-w)$     &$(1-x)^2(1+y)(1+w)(1-w)$\\
0010&$(1-x)^2(1-y)(1+w)(1-w)$     &$(1-x)^2(1+y)(1+w)(1-w)$\\
0100&$(1+x)(1-x)(1-y)(1+w)^2$     &$(1+x)(1-x)(1+y)(1-w)^2$\\
1000&$(1+x)(1-x)(1-y)(1+w)^2$     &$(1+x)(1-x)(1+y)(1-w)^2$\\
0011&$(1+x)(1-x)(1-y)(1+w)(1-w)$  &$(1+x)(1-x)(1+y)(1+w)(1-w)^{\dagger}$\\
0101&$(1-x)^2(1-y)(1-w)^2$       &$(1-x)^2(1+y)(1+w)^{2\ddagger}$\\
1001&$(1+x)(1-x)(1-y)(1+w)(1-w)$  &$(1+x)(1-x)(1+y)(1+w)(1-w)^{\dagger}$\\
    \hline
    \multicolumn{3}{c}{$P(z_j|s_j,\tau_2,t)$}\\
\hline
& \multicolumn{2}{|c|}{$z_j$}\\
    \hline
$s_j$    &00                              &01\\
    \hline
0000&$(1+x)^2(1+y)(1+w)^{2*}$          &$(1+x)(1-x)(1-y)(1+w)(1-w)$\\
0001&$(1+x)^2(1+y)(1+w)(1-w)^{\ddagger}$       &$(1+x)(1-x)(1-y)(1+w)^{2\ddagger}$\\
0010&$(1+x)(1-x)(1+y)(1+w)^{2\ddagger}$       &$(1+x)^2(1-y)(1+w)(1-w)^{\ddagger}$\\
0100&$(1+x)^2(1+y)(1+w)(1-w)^{\ddagger}$       &$(1+x)(1-x)(1-y)(1-w)^2$\\
1000&$(1+x)(1-x)(1+y)(1+w)^{2\ddagger}$       &$(1-x)^2(1-y)(1+w)(1-w)$\\
0011&$(1+x)(1-x)(1+y)(1+w)(1-w)^{\ddagger}$          &$(1+x)^2(1-y)(1+w)^{2\ddagger}$\\
0101&$(1+x)^2(1+y)(1-w)^{2\ddagger}$    &$(1+x)(1-x)(1-y)(1+w)(1-w)^{\ddagger}$\\
1001&$(1+x)(1-x)(1+y)(1+w)(1-w)^{\ddagger}$    &$(1-x)^2(1-y)(1+w)^{2\ddagger}$\\
    \hline
    \hline
&10                           &11\\
    \hline
0000&$(1+x)(1-x)(1-y)(1+w)(1-w)$        &$(1-x)^2(1+y)(1-w)^2$\\
0001&$(1+x)(1-x)(1-y)(1-w)^2$     &$(1-x)^2(1+y)(1+w)(1-w)$\\
0010&$(1-x)^2(1-y)(1+w)(1-w)$     &$(1+x)(1-x)(1+y)(1-w)^2$\\
0100&$(1+x)(1-x)(1-y)(1+w)^{2\ddagger}$     &$(1-x)^2(1+y)(1+w)(1-w)$\\
1000&$(1+x)^2(1-y)(1+w)(1-w)^{\ddagger}$     &$(1+x)(1-x)(1+y)(1-w)^2$\\
0011&$(1-x)^2(1-y)(1-w)^2$       &$(1+x)(1-x)(1+y)(1+w)(1-w)^{\ddagger}$\\
0101&$(1+x)(1-x)(1-y)(1+w)(1-w)^{\ddagger}$  &$(1-x)^2(1+y)(1+w)^{2\ddagger}$\\
1001&$(1+x)^2(1-y)(1-w)^{2\ddagger}$  &$(1+x)(1-x)(1+y)(1+w)(1-w)^{\ddagger}$\\
\hline
\end{tabular}    
\caption{Likelihood calculations for all site patterns and internal states of Farris topology.
Maxima determined row-wise (i.e., by site pattern).
All values multiplied by $1/32$.
Key: $^*$ unique maximum value corresponding to unique internal state; $^\dagger$ unique maximum value corresponding to multiple internal states; $^\ddagger$ multiple maximum values corresponding to multiple internal states.}
\label{tab:likelihoods}
\end{table}

@@

To simplify matters, we will first focus on bounding the the terms with the generating probabilities as they are more complicated.
Let
\begin{align}
\label{eq:information}
H(\mathbf{p};x, y, w) &= \sum_{j} P(s_j | \tau^*, \{x^*,y^*\}) \cdot \log P(s_j | \tau', \{x,y,w\}) \nonumber \\ 
&:= \sum_{j} p(s_j) \cdot \log P(s_j | \tau', \{x,y,w\}).
\end{align}
By Gibbs inequality,
$$
H(\mathbf{p};x, y, w) \le H(\mathbf{p};x^*,y^*, y^*),
$$
though this bound can be made tighter.
In our simplifications we will make use of the constants
$$
a = p(0100) + p(1000) + p(0011) + p(1001) = \left\{
	\begin{array}{ll}
		(4-2x^{*2}-2y^{*2})/8  & \mbox{Farris zone,} \\
		(4-2x^{*2}y^{*}-2y^{*3})/8  & \mbox{Felsenstein zone,}
	\end{array}
\right.
$$
$$
b = p(0101) = \left\{
	\begin{array}{ll}
		(1+x^{*2}+y^{*2}-4x^{*}y^{*2}+x^{*2}y^{*2})/8  & \mbox{Farris zone,} \\
		(1-2x^{*}y^{*}-2x^{*}y^{*2}+x^{*2}y^{*}+y^{*3}+x^{*2}y^{*2})/8  & \mbox{Felsenstein zone,}
	\end{array}
\right.
$$
$$
c = p(0001) + p(0010) + p(0100) + p(1000) = (4-4x^{*2}y^{*2})/8
$$
and
$$
d = p(0101) + p(1001) = \left\{
	\begin{array}{ll}
		(2-4x^{*}y^{*2}+2x^{*2}y^{*2})/8  & \mbox{Farris zone} \\
		(2-4x^{*}y^{*}+2x^{*2}y^{*2})/8  & \mbox{Felsenstein zone}
	\end{array}
\right.
$$
with corresponding bounds
$$
0 \le a,c \le \frac{1}{2}
$$
and
$$
0 \le b,d \le \frac{1}{4}.
$$

Our key approach will be to decompose the likelihood as
\begin{equation}
\label{eq:partial-likelihood}
L^{(i)}(x, y, w) = H(\mathbf{p}; x, y, w) + L^{p,(i)}(x,y,w)
\end{equation}
where the first term is \eqref{eq:information} and the second term will be a weighted sum of the $\log$s of values from Table~\ref{tab:likelihoods}, the superscript indicating which choices for ancestral state likelihood values we choose.
Define
$$
\{\hat{x}, \hat{y}, \hat{w}, \hat{i}\} = \arg\max_{\{x,y,w,i\}} \ L^{p,(i)}(x,y,w).
$$
We have upper and lower bounds for \eqref{eq:partial-likelihood} in the form
$$
H(\mathbf{p}; \hat{x}, \hat{y}, \hat{w}) + L^{p,(\hat{i})}(\hat{x}, \hat{y}, \hat{w}) \le \max_{x,y,w} \ L^{(i)}(x,y,w) \le H(\mathbf{p}; x^{*}, y^{*}, y^{*}) + L^{p,(\hat{i})}(\hat{x}, \hat{y}, \hat{w}).
$$

@@

The likelihood for the Farris zone topology is less complicated than that of the Felsenstein zone since only one site pattern has an ambiguous likelihood---all site patterns except \texttt{0101} have a fixed value for their likelihood when taking the maximum.
We decompose the full likelihood

where $L_{fa}^{(i)}(x,y,w)$ takes one of three forms:
$$
L_{fa}^{(1)} = (2-a)\log(1+x)+a\log(1-x)+(2-a-2b)\log(1+w)+(a+2b)\log(1-w)+\log(1+y),
$$
$$
L_{fa}^{(2)} = (2-a-2b)\log(1+x)+(a+2b)\log(1-x)+(2-a)\log(1+w)+a\log(1-w)+\log(1+y),
$$
$$
L_{fa}^{(3)} = (2-a)\log(1+x)+a\log(1-x)+(2-a)\log(1+w)+a\log(1-w)+(1-b)\log(1+y)+b\log(1-y).
$$
These likelihoods have maxima
$$
\hat{L}_{fa}^{(1)} = \hat{L}_{fa}^{(2)} = (2-a)\log(2-a)+a\log(a)+(2-a-2b)\log(2-a-2b)+(a+2b)\log(a+2b)+\log(2),
$$
$$
\hat{L}_{fa}^{(3)} = (2-a)\log(2-a)+a\log(a)+(2-a)\log(2-a)+a\log(a)+(1-b)\log(1-b)+b\log(b)+\log(2).
$$

\textbf{TODO}: cases for $\{x^{*}, y^{*}\}$ where each of the above is the corresponding maximum.
I think the first is greater than or equal to the second in most cases, though not all (i.e., when $b$ is large but $a$ is small).

@@

We construct similar bounds on the Felsenstein likelihood, though to make matters simpler we replace all $(1+y)$ terms in Table~\ref{tab:likelihoods} with $(1-y)$ to obtain a lower bound, performing the reverse substitution to obtain upper bounds.
As in \eqref{eq:partial-likelihood} we see that the likelihood in this case has one of six forms:
$$
L_{fe}^{(1)} = 2\log(1+x)+(2-c-2d)\log(1+w)+(c+2d)\log(1-w)+\log(1-y),
$$
$$
L_{fe}^{(2)} = (2-c-2d)\log(1+x)+(c+2d)\log(1-x)+2\log(1+w)+\log(1-y),
$$
$$
L_{fe}^{(3)} = (2-2d)\log(1+x)+2d\log(1-x)+(2-c)\log(1+w)+c\log(1-w)+\log(1-y),
$$
$$
L_{fe}^{(4)} = (2-c)\log(1+x)+c\log(1-x)+(2-2d)\log(1+w)+2d\log(1-w)+\log(1-y),
$$
$$
L_{fe}^{(5)} = (2-d)\log(1+x)+d\log(1-x)+(2-c-d)\log(1+w)+(c+d)\log(1-w)+\log(1-y)
$$
and
$$
L_{fe}^{(6)} = (2-c-d)\log(1+x)+(c+d)\log(1-x)+(2-d)\log(1+w)+d\log(1-w)+\log(1-y)
$$
with maxima
$$
\hat{L}_{fe}^{(1)} = \hat{L}_{fe}^{(2)} = 2\log(2)+(2-c-2d)\log(2-c-2d)+(c+2d)\log(c+2d)
$$
$$
\hat{L}_{fe}^{(3)} = \hat{L}_{fe}^{(4)} = (2-2d)\log(2-2d)+2d\log(2d)+(2-c)\log(2-c)+c\log(c)
$$
and
$$
\hat{L}_{fe}^{(5)} = \hat{L}_{fe}^{(6)} = (2-d)\log(2-d)+d\log(d)+(2-c-d)\log(2-c-d)+(c+d)\log(c+d).
$$

\textbf{TODO}: cases for $\{x^{*}, y^{*}\}$ where each of the above is the corresponding maximum.
I think the first is always greater than or equal to the other two, but I haven't been able to show it yet.
It may take a little information theory--fu, or just standard application of the bounds in the introduction to this section.

@@

Set $\{x^*, y^*\} = \{0.5, 0.5\}$.
For a Farris zone generating topology, we have
$$
a =  0.375, \ b =  0.1328125,
$$
resulting in
$$
L_{fa}^{(1)} = L_{fa}^{(2)} \approx 1.24637119863,
$$
$$
L_{fa}^{(3)} \approx 1.14372724688
$$
and
$$
H_{fa}(\mathbf{p};x^*,y^*, y^*) \approx -1.99591752302.
$$
Thus computing the upper bound yields
$$
\max_{x,y,w} \ L_{fa}(x,y,w) \le -0.749546324382
$$
and it remains to show that 
$$
\max_{x,y,w} \ L_{fe}(x,y,w) \ge -0.749546324382.
$$

For the same generating parameters,
$$
c = 0.46875, \ d = 0.203125,
$$
resulting in
$$
L_{fe}^{(1)} = L_{fe}^{(2)} \approx 1.40196030769, 
$$
$$
L_{fe}^{(3)} = L_{fe}^{(4)} \approx 0.674162530673
$$
and
$$
L_{fe}^{(5)} = L_2^{(6)} \approx 0.838975450706.
$$
Calculating shows
$$
H_{fe}(\mathbf{p}; \hat{x}, \hat{y}, \hat{w}) = -2.06377559511,
$$
and
$$
\max_{x,y,w} \ L_{fe}(x,y,w) \ge -0.661815287426 > -0.749546324382 \ge \max_{x,y,w} \ L_{fa}(x,y,w)
$$
where site patterns are generated from the Farris zone topology.
Wait, really?
Did I just show something???

>>>>

%% Useful packages
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}

\graphicspath{ {figures/} }

@@

Define the Farris topology $\tau_1$ and the Felsenstein topology $\tau_2$ as in Figure~\ref{fig:farris-fels-top}.

\begin{figure}
\centering
\begin{subfigure}{.45\linewidth}
\centering
\includegraphics[width=.95\textwidth]{farris_blank}
\caption[short]{Farris topology $\tau_1$}
\end{subfigure}
\begin{subfigure}{.45\linewidth}
\centering
\includegraphics[width=.95\textwidth]{felsenstein_blank}
\caption[short]{Felsenstein topology $\tau_2$}
\end{subfigure}
\caption{Two simple topologies}
\end{figure}

Call $t=\{x,y,w\}$ and $t^*=\{x,y,y\}$, i.e., $t^*$ is the case where the bottom three branches all share the same parameter, the classical construction of this topology.
The branch length parameters are such that the probability of a change in character along the top two branches is $p_x=1-2x$, with corresponding equalities for the other two branches.
Here, $\{p_x,p_y,p_z\}\in[0,1/2]^3$ and $\{x,y,z\}\in[0,1]^3$.

@@

\begin{table}
\centering
\begin{tabular}{|l|l|l|}
\multicolumn{3}{c}{Classical}\\
    \hline
$s_j$   &$P(s_j|\tau_1,t^*)$&$P(s_j|\tau_2,t^*)$\\
    \hline
0000&$1+x^2+y^2+4xy^2+x^2y^2$&$1+2xy+2xy^2+x^2y+y^3+x^2y^2$\\
0001&$1+x^2-y^2-x^2y^2$&$1+x^2y-y^3-x^2y^2$\\
0010&$1-x^2+y^2-x^2y^2$&$1-x^2y+y^3-x^2y^2$\\
0100&$1+x^2-y^2-x^2y^2$&$1+x^2y-y^3-x^2y^2$\\
1000&$1-x^2+y^2-x^2y^2$&$1-x^2y+y^3-x^2y^2$\\
0011&$1-x^2-y^2+x^2y^2$&$1+2xy-2xy^2-x^2y-y^3+x^2y^2$\\
0101&$1+x^2+y^2-4xy^2+x^2y^2$&$1-2xy-2xy^2+x^2y+y^3+x^2y^2$\\
1001&$1-x^2-y^2+x^2y^2$&$1-2xy+2xy^2-x^2y-y^3+x^2y^2$\\
    \hline
\multicolumn{3}{c}{Modified}\\
    \hline
$s_j$   &$P(s_j|\tau_1,t)$&$P(s_j|\tau_2,t)$\\
    \hline
0000&$1+x^2+y^2+4xyw+x^2y^2$&$1+2xy+2xyw+x^2w+y^2w+x^2y^2$\\
0001&$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
0010&$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
0100&$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
1000&$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
0011&$1-x^2-y^2+x^2y^2$&$1+2xy-2xyw-x^2w-y^2w+x^2y^2$\\
0101&$1+x^2+y^2-4xyw+x^2y^2$&$1-2xy-2xyw+x^2w+y^2w+x^2y^2$\\
1001&$1-x^2-y^2+x^2y^2$&$1-2xy+2xyw-x^2w-y^2w+x^2y^2$\\
    \hline
\end{tabular}    
\caption{Site pattern probabilities.
All values are multiplied by $1/8$.}
\label{tab:sitepatprob}
\end{table}

\begin{table}
\centering
\begin{tabular}{|l|ll|}
\multicolumn{3}{c}{$P(z_j|s_j,\tau_1,t)$}\\
\hline
& \multicolumn{2}{|c|}{$z_j$}\\
    \hline
$s_j$    &00                              &01\\
    \hline
0000&$(1+x)^2   (1+w)(1+y)^{2*}$          &$(1+x)^2   (1-w)(1-y)^2$\\
0001&$(1+x)^2   (1+w)(1+y)(1-y)^*$        &$(1+x)^2   (1-w)(1+y)(1-y)$\\
0010&$(1+x)^2   (1+w)(1+y)(1-y)^*$        &$(1+x)^2   (1-w)(1+y)(1-y)$\\
0100&$(1+x)(1-x)(1+w)(1+y)^{2*}$          &$(1+x)(1-x)(1-w)(1-y)^2$\\
1000&$(1+x)(1-x)(1+w)(1+y)^{2*}$          &$(1+x)(1-x)(1-w)(1-y)^2$\\
0011&$(1+x)(1-x)(1+w)(1+y)(1-y)^{\dagger}$&$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
0101&$(1+x)^2   (1+w)(1-y)^{2\ddagger}$   &$(1+x)^2   (1-w)(1+y)^{2\ddagger}$\\
1001&$(1+x)(1-x)(1+w)(1+y)(1-y)^{\dagger}$&$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
    \hline
    \hline
&10                           &11\\
    \hline
0000&$(1-x)^2   (1-w)(1+y)^2$     &$(1-x)^2   (1+w)(1-y)^2$\\
0001&$(1-x)^2   (1-w)(1+y)(1-y)$  &$(1-x)^2   (1+w)(1+y)(1-y)$\\
0010&$(1-x)^2   (1-w)(1+y)(1-y)$  &$(1-x)^2   (1+w)(1+y)(1-y)$\\
0100&$(1+x)(1-x)(1-w)(1+y)^2$     &$(1+x)(1-x)(1+w)(1-y)^2$\\
1000&$(1+x)(1-x)(1-w)(1+y)^2$     &$(1+x)(1-x)(1+w)(1-y)^2$\\
0011&$(1+x)(1-x)(1-w)(1+y)(1-y)$  &$(1+x)(1-x)(1+w)(1+y)(1-y)^{\dagger}$\\
0101&$(1-x)^2   (1-w)(1-y)^2$     &$(1-x)^2   (1+w)(1+y)^{2\ddagger}$\\
1001&$(1+x)(1-x)(1-w)(1+y)(1-y)$  &$(1+x)(1-x)(1+w)(1+y)(1-y)^{\dagger}$\\
    \hline
    \multicolumn{3}{c}{$P(z_j|s_j,\tau_2,t)$}\\
\hline
& \multicolumn{2}{|c|}{$z_j$}\\
    \hline
$s_j$    &00                              &01\\
    \hline
0000&$(1+x)^2   (1+w)(1+y)^{2*}$           &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
0001&$(1+x)^2   (1+w)(1+y)(1-y)^{\ddagger}$&$(1+x)(1-x)(1-w)(1+y)^{2\ddagger}$\\
0010&$(1+x)(1-x)(1+w)(1+y)^{2\ddagger}$    &$(1+x)^2   (1-w)(1+y)(1-y)^{\ddagger}$\\
0100&$(1+x)^2   (1+w)(1+y)(1-y)^{\ddagger}$&$(1+x)(1-x)(1-w)(1-y)^2$\\
1000&$(1+x)(1-x)(1+w)(1+y)^{2\ddagger}$    &$(1-x)^2   (1-w)(1+y)(1-y)$\\
0011&$(1+x)(1-x)(1+w)(1+y)(1-y)^{\ddagger}$&$(1+x)^2   (1-w)(1+y)^{2\ddagger}$\\
0101&$(1+x)^2   (1+w)(1-y)^{2\ddagger}$    &$(1+x)(1-x)(1-w)(1+y)(1-y)^{\ddagger}$\\
1001&$(1+x)(1-x)(1+w)(1+y)(1-y)^{\ddagger}$&$(1-x)^2   (1-w)(1+y)^{2\ddagger}$\\
    \hline
    \hline
&10                           &11\\
    \hline
0000&$(1+x)(1-x)(1-w)(1+y)(1-y)$             &$(1-x)^2   (1+w)(1-y)^2$\\
0001&$(1+x)(1-x)(1-w)(1-y)^2$                &$(1-x)^2   (1+w)(1+y)(1-y)$\\
0010&$(1-x)^2   (1-w)(1+y)(1-y)$             &$(1+x)(1-x)(1+w)(1-y)^2$\\
0100&$(1+x)(1-x)(1-w)(1+y)^{2\ddagger}$      &$(1-x)^2   (1+w)(1+y)(1-y)$\\
1000&$(1+x)^2   (1-w)(1+y)(1-y)^{\ddagger}$  &$(1+x)(1-x)(1+w)(1-y)^2$\\
0011&$(1-x)^2   (1-w)(1-y)^2$                &$(1+x)(1-x)(1+w)(1+y)(1-y)^{\ddagger}$\\
0101&$(1+x)(1-x)(1-w)(1+y)(1-y)^{\ddagger}$  &$(1-x)^2   (1+w)(1+y)^{2\ddagger}$\\
1001&$(1+x)^2   (1-w)(1-y)^{2\ddagger}$      &$(1+x)(1-x)(1+w)(1+y)(1-y)^{\ddagger}$\\
\hline
\end{tabular}    
\caption{Likelihood calculations for all site patterns and internal states of Farris topology.
Maxima determined row-wise (i.e., by site pattern).
All values multiplied by $1/32$.
Key: $^*$ unique maximum value corresponding to unique internal state; $^\dagger$ unique maximum value corresponding to multiple internal states; $^\ddagger$ multiple maximum values corresponding to multiple internal states.}
\label{tab:likelihoods}
\end{table}

@@

To simplify matters, we will first focus on bounding the the terms with the generating probabilities as they are more complicated.
Assume $\tau^*=\tau_1$ and $t^*=\{x,y,y\}$ so that
\begin{align}
\label{eq:information}
H_{\tau_1}(x, y;x', y', w') &= \sum_{j} P(s_j | \tau_1, \{x,y,y\}) \cdot \log P(s_j | \tau', \{x',y',w'\}) \nonumber \\ 
&:= \sum_{j} p(s_j) \cdot \log P(s_j | \tau', \{x',y',w'\}).
\end{align}
By Gibbs inequality,
$$
H_{\tau_1}(x, y;x', y', w') \le H_{\tau_1}(x, y;x, y, y),
$$
though this bound can be made tighter.
In our simplifications we will make use of the constants
$$
a = p(0100) + p(1000) + p(0011) + p(1001) = (4-2x^2-2y^2)/8,
$$
$$
b = p(1000) + p(0010) + p(0011) + p(1001) = (4-2x^2-2y^2)/8,
$$
$$
c = p(0001) + p(0010) + p(0100) + p(1000) = (4-4x^2y^2)/8
$$
and
$$
d = p(0101) + p(1001) = (2-4xy^2+2x^2y^2)/8
$$
with corresponding bounds
$$
0 \le a,b,c \le \frac{1}{2}
$$
and
$$
0 \le d \le \frac{1}{4}.
$$

Our key approach will be to decompose the likelihood as
\begin{equation}
\label{eq:partial-likelihood}
L^{(i)}(x', y', w') = H_{\tau_1}(x,y;x',y',w') + L^{p,(i)}(x',y',w')
\end{equation}
where the first term is \eqref{eq:information} and the second term will be a weighted sum of the $\log$s of values from Table~\ref{tab:likelihoods}, the superscript indicating which choices for ancestral state likelihood values we choose.
Define
$$
\{\hat{x}, \hat{y}, \hat{w}, \hat{i}\} = \arg\max_{\{x,y,w,i\}} \ L^{p,(i)}(x,y,w).
$$
We have upper and lower bounds for \eqref{eq:partial-likelihood} in the form
$$
H_{\tau_1}(x,y; \hat{x}, \hat{y}, \hat{w}) + L^{p,(\hat{i})}(\hat{x}, \hat{y}, \hat{w}) \le \max_{x,y,w} \ L^{(i)}(x,y,w) \le H(x,y;x,y,y) + L^{p,(\hat{i})}(\hat{x}, \hat{y}, \hat{w}).
$$

@@

The likelihood for the Farris topology is less complicated than that of the Felsenstein since only one site pattern has an ambiguous likelihood---all site patterns except \texttt{0101} have a fixed value for their likelihood when taking the maximum.
To fix this value, we will consider an upper bound on the partial likelihood by setting the value for site pattern 0011 to $(1+x)^2(1+w)(1+y)^2$.
Here the partial likelihood $L_{fa}^{p,(i)}(x,y,w)$ takes the form:
$$
L_{fa}^{p} = (2-a)\log(1+x)+a\log(1-x)+(2-b)\log(1+y)+b\log(1-y)+\log(1+w),
$$
and this has a maximum
$$
\hat{L}_{fa}^{p} = (2-a)\log(2-a)+a\log(a)+(2-b)\log(2-b)+b\log(b)+\log(2).
$$

@@

We construct similar bounds on the Felsenstein likelihood, though to make matters simpler we replace all $(1+y)$ terms in Table~\ref{tab:likelihoods} with $(1-y)$ to obtain a lower bound.
Unlike the Farris partial likelihood, the likelihood in this case has one of six forms:
$$
L_{fe}^{(1)} = 2\log(1+x)+(2-c-2d)\log(1+y)+(c+2d)\log(1-y)+\log(1-w),
$$
$$
L_{fe}^{(2)} = (2-c-2d)\log(1+x)+(c+2d)\log(1-x)+2\log(1+y)+\log(1-w),
$$
$$
L_{fe}^{(3)} = (2-2d)\log(1+x)+2d\log(1-x)+(2-c)\log(1+y)+c\log(1-y)+\log(1-w),
$$
$$
L_{fe}^{(4)} = (2-c)\log(1+x)+c\log(1-x)+(2-2d)\log(1+y)+2d\log(1-y)+\log(1-w),
$$
$$
L_{fe}^{(5)} = (2-d)\log(1+x)+d\log(1-x)+(2-c-d)\log(1+y)+(c+d)\log(1-y)+\log(1-w)
$$
and
$$
L_{fe}^{(6)} = (2-c-d)\log(1+x)+(c+d)\log(1-x)+(2-d)\log(1+y)+d\log(1-y)+\log(1-w)
$$
with maxima
$$
\hat{L}_{fe}^{(1)} = \hat{L}_{fe}^{(2)} = 2\log(2)+(2-c-2d)\log(2-c-2d)+(c+2d)\log(c+2d)
$$
$$
\hat{L}_{fe}^{(3)} = \hat{L}_{fe}^{(4)} = (2-2d)\log(2-2d)+2d\log(2d)+(2-c)\log(2-c)+c\log(c)
$$
and
$$
\hat{L}_{fe}^{(5)} = \hat{L}_{fe}^{(6)} = (2-d)\log(2-d)+d\log(d)+(2-c-d)\log(2-c-d)+(c+d)\log(c+d).
$$

\textbf{TODO}: cases for $\{x, y\}$ where each of the above is the corresponding maximum.
I think the first is always greater than or equal to the other two, but I haven't been able to show it yet.
It may take a little information theory--fu, or just standard application of the bounds in the introduction to this section.

@@

Set $\{x, y\} = \{0.25, 0.25\}$.
For a Farris zone generating topology, we have
$$
a = b = 0.46875
$$
resulting in
$$
\hat{L}_{fa}^{p}(\hat{x}, \hat{y}, \hat{w}) \approx -2.17803560694
$$
and
$$
H_{fa}(x,y;x,y,y) \approx -2.07509999739.
$$
Computing the upper bound yields
$$
\max_{x,y,w} \ L_{fa}(x,y,w) \le -4.25313560433
$$
and it remains to show that 
$$
\max_{x,y,w} \ L_{fe}(x,y,w) \ge -4.25313560433
$$

For the same generating parameters,
$$
c = 0.498046875, \ d = 0.2431640625,
$$
resulting in
$$
\hat{L}_{fe}(\hat{x}, \hat{y}, \hat{w}) \approx -2.07919739112
$$
Calculating shows%
    %
    \footnote{It's not a typo that $\hat{L}_{fe}$ and $H_{fe}(x,y; \hat{x}, \hat{y}, \hat{w})$ are the same---at least not in the code.
        I should try to show why this is the case.
        }
    %
$$
H_{fe}(x,y; \hat{x}, \hat{y}, \hat{w}) \approx -2.07919739112
$$
and
$$
\max_{x,y,w} \ L_{fe}(x,y,w) \ge -4.15839478224 > -4.25313560433 \ge \max_{x,y,w} \ L_{fa}(x,y,w)
$$
where site patterns are generated from the Farris zone topology.
Confirming, run \texttt{python}'s numerical optimization solver for $\{.25, .25\}$.
Since we are computing a constrained maximization, we can find a global maximum using basin hopping, and doing so yields
$$
\max_{x,y,w} \ L_{fa}(x,y,w) \approx -4.5101771294218045
$$
and
$$
\max_{x,y,w} \ L_{fe}(x,y,w) \approx -4.1544859851260281.
$$
Woohoo!

\paragraph{General example}

We want to find values for $\{x,y\}$ where
$$
H_{fe}(x,y; 1, 0, 1-c-2d) + \hat{L}_{fe}(c,d) \ge H_{fa}(x,y; x, y, y) + \hat{L}_{fa}(a,b).
$$
We can cheat a bit and plot the values at which this is true.
This may give us some insight into what form these regions of inconsistency take---see Fig.~\ref{fig:inconsistency-farris}.
